---
title: "DEGA_processed_data"
author: "Ben Barry"
date: "2025-02-04"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r check-install}
#create a vector of required packages and compare with installed packages for efficient installation 

pkg <- c("SummarizedExperiment", "extraChIPs", "edgeR", "tidyverse")
not_inst <- setdiff(pkg, rownames(installed.packages()))


# if there are packages required which arent installed, install them. 

if (length(not_inst)) BiocManager::install(not_inst)
```


```{r load-libraries}
library(tidyverse)
library(SummarizedExperiment)
library(edgeR)
library(magrittr)
library(patchwork)
library(extraChIPs)
```


``` {r load-outlier_summarised-exp}
SE_outlier_removed <- read_rds("/Users/benbarry/Documents/Telethon2025/data/processed/SE_outlier_removed.rds")

```



```{r DesignMatrix}

# creating the design matrix for analysis
# this design matrix uses the first column as the intercept term (in this case healthy controls)

model <- model.matrix(~ disease_state, data = colData(SE_outlier_removed))

model

#creating a contrast matrix 
#because the intercept term is defined we are "comparing" with that

contrastMatrix <- makeContrasts(
  T2D_v_DN = disease_statetype2_diabetes - disease_stateDiabetic_nephropathy,  
  Hv_v_T2D = disease_statetype2_diabetes,
  Hv_v_DN = disease_stateDiabetic_nephropathy,
  levels = model)

contrastMatrix

# plot the biological coefficient of variance.

SE_outlier_removed %>%
  calcNormFactors(method = "TMM") %>%
  estimateDisp(design = model) %>%
  plotBCV()

#create a GLM QL fit - this is the recommended approach now. 
#note: important to specify normalization method here.

fit <- SE_outlier_removed %>%
  calcNormFactors(method = "TMM") %>%
  estimateDisp(design = model) %>%
  glmQLFit(model)


```

```{r QLFTests Hv-vs-T2D}

# GLM-QL test for DE comparing Healthy vs Diabetes. 

Hv_v_T2D <- glmQLFTest(fit, contrast = contrastMatrix[, "Hv_v_T2D"]) %>%
  topTags(n = Inf) %>%
  pluck("table") %>%      #pluck() pulls "table" out of the topTags obj
  as_tibble() %>%
  mutate(Significant = PValue < 0.05)

# return the number of significant and NS genes.

Hv_v_T2D %>%
  summarise(
    n = dplyr::n(),
    .by = Significant)


# plot an MA plot (logCPM vs logFC) coloured by significance. 

Hv_v_T2D %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  geom_point(aes(colour = Significant)) +
  scale_colour_manual(values = c("black", "red"))
    


# GLMTREAT test for DE comparing Healthy vs Diabetes. 

Hv_v_T2D_treat <- glmTreat(
  fit, contrast = contrastMatrix[, "Hv_v_T2D"],  lfc = log2(1.2)) %>%
  topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05)


# plot an MA plot (logCPM vs logFC) coloured by significance. 

Hv_v_T2D_treat %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  geom_point(aes(colour = Significant)) +
  geom_smooth(method = 'lm', se = FALSE) +
  scale_colour_manual(values = c("black", "red"))


# return the number of significant and NS genes.  

Hv_v_T2D_treat %>%
  summarise(n = dplyr::n(),
            .by = Significant)



```

```{r HV-vs-DN}

# GLM-QL test for DE comparing Healthy vs Diabetic Nepropathy

Hv_v_DN <- glmQLFTest(fit, contrast = contrastMatrix[, "Hv_v_DN"]) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05)


# plot an MA plot (logCPM vs logFC) coloured by significance. 

Hv_v_DN %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  geom_point(aes(colour = Significant)) +
  scale_colour_manual(values = c("black", "red"))


# return the number of significant and NS genes.    

Hv_v_DN %>%
  summarise(n = dplyr::n(),
            .by = Significant)



#GLMTreat test for DE comparing Healthy vs Diabetic Nepropathy

Hv_v_DN_treat <- glmTreat(fit, contrast = contrastMatrix[, "Hv_v_DN"],  lfc = log2(1.5)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05)


# plot an MA plot (logCPM vs logFC) coloured by significance.

Hv_v_DN_treat%>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  geom_point(aes(colour = Significant)) +
  geom_smooth(method = 'lm', se = FALSE) +
  scale_colour_manual(values = c("black", "red"))
 

# return the number of significant and NS genes.    

Hv_v_DN_treat%>%
  summarise(n = dplyr::n(),
            .by = Significant)


```

```{r T2D-vs-DN}
#GLM-QL test for DE comparing Diabetes with Diabetic Neropathy

T2D_v_DN <- glmQLFTest(fit, contrast = contrastMatrix[, "T2D_v_DN"]) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05)


# plot an MA plot (logCPM vs logFC) coloured by significance. 

T2D_v_DN %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  geom_point(aes(colour = Significant)) +
  scale_colour_manual(values = c("black", "red"))


# return the number of significant and NS genes.  

T2D_v_DN %>%
  summarise(n = dplyr::n(),
            .by = Significant)



#GLMTreat for DE comparing Diabetes with Diabetic Neropathy

T2D_v_DN_treat <- glmTreat(fit, contrast = contrastMatrix[, "T2D_v_DN"], lfc = log2(1.5)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05)


# plot an MA plot (logCPM vs logFC) coloured by significance. 

T2D_v_DN_treat%>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  geom_point(aes(colour = Significant)) +
  geom_smooth(method = 'lm', se = FALSE) +
  scale_colour_manual(values = c("black", "red"))


# return the number of significant and NS genes. 

T2D_v_DN_treat%>%
  summarise(n = dplyr::n(),
            .by = Significant)


```
# RUVg normalisation.

"If no genes are known a priori not to be influenced by the covariates of interest, one can obtain a set of “in-silico empirical” negative controls, e.g., least significantly DE genes based on a first-pass DE analysis performed prior to RUVg normalization."

the TMM normalization above is okay, however there is still some variance indicated by the slope of the lm. 
here we are trying RUVg. 


```{r ruvg-nomralisation}
library(RUVSeq)
#step 1: run a normal DGE 
  #1.1: same model design as before.

model <- model.matrix(~ disease_state, data = colData(SE_outlier_removed))
model


# 1.2: same fit as before using TMM for normlaisation

fit <- SE_outlier_removed %>%
  calcNormFactors(method = "TMM") %>%
  estimateDisp(design = model) %>%
  glmQLFit(model)


# 1.3: perform GLMQLF test.
  # note: here we have no contrasts and are instead performing and ANOVA-like test

QLFTest <- glmQLFTest(fit, coef = 2:3)
QLFTest


# step 2: return the differential expressed genes.

QLFtop <- QLFTest %>%
  topTags(n = Inf) %>%
  pluck("table") %>%
  dplyr::select(
    starts_with("gene"),
    logCPM,
    'F',
    PValue,
    FDR)


#step 3: create a set of 5000 of genes with the highest PValue (ie. not changing). 
  # note: this is somewhat arbitrary

emperical <- QLFtop %>%
  dplyr::arrange(desc(PValue)) %>%
  .[1:5000,] %>%
  .$gene_id


# step 4: run RUVg (where estimated factor of unwanted variation (k) = 1)

ruvg <- RUVg(assay(SE_outlier_removed, "counts"), emperical, k = 1) 
ruvg$W

# due to compatibility issues  - convert to a DGElist object 
ruv_dge <- DGEList(
  counts = ruvg$normalizedCounts,
  samples = cbind(colData(SE_outlier_removed), ruvg$W),
  genes = rowRanges(SE_outlier_removed))



```

```{r RUVG_fit_and_model}

# create a new design model including the estimated variation.

RUVg_model <- model.matrix(~ disease_state + W_1, data = ruv_dge$samples)


# create a new fit
  # note: We still use TMM normalisation here 

RUVg_fit <- ruv_dge %>%
  calcNormFactors(method = "TMM") %>%
  estimateDisp(RUVg_model) %>%
  glmQLFit(RUVg_model)

# create a new contrast matrix with the levels from model. 

RUVg_contrasts <- makeContrasts(
  T2D_v_DN = disease_statetype2_diabetes - disease_stateDiabetic_nephropathy,  
  Hv_v_T2D = disease_statetype2_diabetes,
  Hv_v_DN = disease_stateDiabetic_nephropathy,
  levels = RUVg_model)


# add RUVg normalized counts to se object as an assay for ease of visualization
assay(SE_outlier_removed, "RUVGnorm") <- ruv_dge %>%
  cpm(log=TRUE)
write_rds(SE_outlier_removed, "/Users/benbarry/Documents/Telethon2025/data/processed/SE_outlier_removed.rds", compress = "gz")

# use extraChIPs to plot data 
plotAssayRle(SE_outlier_removed, "RUVGnorm", fill = "disease_state")

plotAssayPCA(
  SE_outlier_removed, "RUVGnorm",
  colour =  "disease_state",
  label = "sampleName")
```

```{r ruvg_normalised_genes}
RUV_T2D_V_DN <- glmTreat(RUVg_fit, contrast = RUVg_contrasts[, "T2D_v_DN"],  lfc = log2(1.5)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05) 
  

RUV_Hv_V_DN <- glmTreat(RUVg_fit, contrast = RUVg_contrasts[, "Hv_v_DN"],  lfc = log2(1.5)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05)

RUV_Hv_V_T2D <- glmTreat(RUVg_fit, contrast = RUVg_contrasts[, "Hv_v_T2D"],  lfc = log2(1.5)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05) 

```



```{r}
RUV_Hv_V_T2D %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
    )
  ) + 
  geom_point(aes(colour = Significant)) +
  geom_smooth(method = 'lm', se = FALSE) +
  scale_colour_manual(values = c("black", "red"))



```


the next set of analysis will take place using new markdown where DGE results are imported. 

```{r saving_data}
write_tsv(RUV_T2D_V_DN, "/Users/benbarry/Documents/Telethon2025/data/processed/RUV_T2D_V_DN.tsv")
write_tsv(RUV_Hv_V_DN, "/Users/benbarry/Documents/Telethon2025/data/processed/RUV_Hv_V_DN.tsv")
write_tsv(RUV_Hv_V_T2D, "/Users/benbarry/Documents/Telethon2025/data/processed/RUV_Hv_V_T2D.tsv")
```


