---
title: "GSEA"
author: "Ben Barry"
date: "2025-04-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load-libraries}
library(tidyverse)
library(dplyr)
library(magrittr)
library(here)
library(clusterProfiler)
library(purrr)
library(reactable)
library(readr)
library(msigdbr)
library(reactable)
library(htmltools)
```

```{r load-data}
RUV_Hv_V_DN <- read_tsv(here("data/processed", "RUV_Hv_V_DN.tsv"))

RUV_Hv_V_T2D <- read_tsv(here("data/processed", "RUV_Hv_V_T2D.tsv"))

RUV_T2D_V_DN <- read_tsv(here("data/processed", "RUV_T2D_V_DN.tsv"))
```

## Gene Set Enrichment Analysis (GSEA)
Similar to ORA, GSEA determines whether predefined sets of genes (e.g., pathways or functional categories) show statistically significant, coordinated differences between two samples. Unlike ORA which uses a user-defined list (eg. significantly DE), GSEA analyzes the entire ranked list of genes to detect subtle but consistent expression changes across a gene set.
For this analysis we will use the same MSIGBD collections

```{r msigdbr}

msigdbr_df <- msigdbr(species = "Homo sapiens") %>%
  dplyr::filter(
      gs_cat == "H" | gs_subcat %in% c("CP:KEGG", "CP:REACTOME", "GO:BP, GO:MF") #, "IMMUNESIGDB"
  )

```

## Healthy control Vs. Type 2 diabetes. 

```{r GSEA-Hv_V_T2D}

geneRanks <- RUV_Hv_V_T2D %>%
  mutate(
    ensemble_id = str_remove(gene_id, "\\.[0-9]+$"),
    signed = setNames(-1 * sign(logFC) * log10(PValue), ensemble_id)
    ) %>%
  arrange(desc(signed)) %>% 
  pluck("signed")

terms <- msigdbr_df %>%
  dplyr::select(gs_name, ensembl_gene)

x <- GSEA(geneRanks,
     TERM2GENE = terms,
     pAdjustMethod = "bonferroni"
     )


cp <- htmltools::tags$caption(
  "Healthy control Vs. Type 2 diabetes."
)

tbl_ID <- "GSEA-Hv-V-T2D"

GSEA_tbl <- x %>%
  pluck("result") %>%
  dplyr::select(enrichmentScore, NES, p.adjust) %>%
  reactable(
    theme = reactableTheme(style = list(fontSize = 12)),
    elementId = tbl_ID
  )

browsable(
  tagList(
    tags$button(
      tagList(fontawesome::fa("download"), "Download as CSV"),
      onclick = glue::glue("Reactable.downloadDataCSV('{tbl_ID}', '{tbl_ID}.csv')")
    ),
    tags$br(), cp,
    GSEA_tbl
  )
)
```




## Healthy control Vs. Diabetic Nephropathy. 

```{r GSEA-Hv_V_DN}

geneRanks <- RUV_Hv_V_DN %>%
  mutate(
    ensemble_id = str_remove(gene_id, "\\.[0-9]+$"),
    signed = setNames(-1 * sign(logFC) * log10(PValue), ensemble_id)
    ) %>%
  arrange(desc(signed)) %>% 
  pluck("signed")

terms <- msigdbr_df %>%
  dplyr::select(gs_name, ensembl_gene)

x <- GSEA(geneRanks,
     TERM2GENE = terms,
     pAdjustMethod = "bonferroni"
     )


cp <- htmltools::tags$caption(
  "Healthy control Vs. Diabetic Nephropathy"
)

tbl_ID <- "GSEA-Hv-V-DN"

GSEA_tbl <- x %>%
  pluck("result") %>%
  dplyr::select(enrichmentScore, NES, p.adjust) %>%
  reactable(
    theme = reactableTheme(style = list(fontSize = 12)),
    elementId = tbl_ID
  )

browsable(
  tagList(
    tags$button(
      tagList(fontawesome::fa("download"), "Download as CSV"),
      onclick = glue::glue("Reactable.downloadDataCSV('{tbl_ID}', '{tbl_ID}.csv')")
    ),
    tags$br(), cp,
    GSEA_tbl
  )
)
```


## Type 2 Diabetes Vs. Diabetic Nephropathy. 

```{r GSEA-T2D_V_DN}

geneRanks <- RUV_T2D_V_DN%>%
  mutate(
    ensemble_id = str_remove(gene_id, "\\.[0-9]+$"),
    signed = setNames(-1 * sign(logFC) * log10(PValue), ensemble_id)
    ) %>%
  arrange(desc(signed)) %>% 
  pluck("signed")

terms <- msigdbr_df %>%
  dplyr::select(gs_name, ensembl_gene)

x <- GSEA(geneRanks,
     TERM2GENE = terms,
     pAdjustMethod = "bonferroni"
     )


cp <- htmltools::tags$caption(
  "Type 2 Diabetes Vs. Diabetic Nephropathy"
)

tbl_ID <- "GSEA-T2D-V-DN"

GSEA_tbl <- x %>%
  pluck("result") %>%
  dplyr::select(enrichmentScore, NES, p.adjust) %>%
  reactable(
    theme = reactableTheme(style = list(fontSize = 12)),
    elementId = tbl_ID
  )

browsable(
  tagList(
    tags$button(
      tagList(fontawesome::fa("download"), "Download as CSV"),
      onclick = glue::glue("Reactable.downloadDataCSV('{tbl_ID}', '{tbl_ID}.csv')")
    ),
    tags$br(), cp,
    GSEA_tbl
  )
)
```