---
title: "Removal of low quality samples"
author: "Ben Barry"
date: "2025-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load_Libraries}
library(tidyverse)
library(extraChIPs)
library(here)
library(edgeR)
library(patchwork)
```



### Removal of HV_3 and HV_4

As determiend previously, normalisation was not sufficient to remove the variance present in the count data introduced by the HV_3 and HV_4 samples. Hence it was decided to exckude these samples from further analyses.

```{r}
SE_outlier_removed <- read_rds(here::here("data/processed/SE_outlier_removed.rds"))


```


### TMM normalisation
With these sampel exclused we normaliased the count data again and plotted these data using Density, RLE and PCA.

```{r TMMnormalisation}

#calc normalisation factors using TMM
assay(SE_outlier_removed, "TmmNorm") <- SE_outlier_removed %>% 
  calcNormFactors(method = "TMM") %>% 
  cpm(log = TRUE)

RLE_A <- SE_outlier_removed %>% 
  plotAssayRle(assay = "TmmNorm", fill = "disease_state") + 
  theme_bw() + 
  scale_fill_brewer(palette = "Accent")

RLE_B <- SE_outlier_removed%>% 
  plotAssayRle(
    assay = "TmmNorm", fill = "disease_state", by_x = "disease_state") +
  theme_bw() + 
  scale_fill_brewer(palette = "Accent")

RLE_A + RLE_B +
  plot_layout(guides = "collect", axes = "collect") &
  theme(legend.position = "bottom")




PCA_A <- SE_outlier_removed %>% 
  plotAssayPCA(
    assay = "TmmNorm",
    colour =  "disease_state",
    label = "sampleName",
    pc_x = 1,
    pc_y = 2
    ) +
  geom_point(shape = 1,size = 2,colour = "black") + 
  scale_colour_brewer(palette = "Accent") + 
  theme_bw()

PCA_B <- SE_outlier_removed %>% 
  plotAssayPCA(
    assay = "TmmNorm",
    colour =  "disease_state",
    label = "sampleName",
    pc_x = 2,
    pc_y = 3
    ) +
  geom_point(shape = 1,size = 2,colour = "black") + 
  scale_colour_brewer(palette = "Accent") + 
  theme_bw()
  

PCA_A + PCA_B +
  plot_layout(guides = "collect", axes = "collect") &
  theme(legend.position = "bottom")


```


Here we observed that removal of HV_3 and HV_4 prior to TMM normalisation appeares to reduce the varience present in the count data across all samples more effectivly. 

From here we moved on to Differential gene expression analysis using `EdgeR`

