---
title: "RUVg Normalisation"
author: "Ben Barry"
date: "2025-04-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load-Libraries}
library(tidyverse)
library(extraChIPs)
library(here)
library(edgeR)
library(patchwork)
library(reactable)
library(htmltools)
library(plotly)
library(pander)
panderOptions("big.mark", ",")
panderOptions("missing", "")
format_p <- function(x) {
  fmt <- ifelse(x < 0.001, "%.2e", "%.3f")
  sprintf(fmt, x)
}
```


```{r load-data}

SE_outlier_removed <- read_rds(here::here("data/processed/SE_outlier_removed.rds"))

```

## RUVg Normalisation.

The primary aim of RUVg normalisation is to identify and a priori set of genes which do not change much between samples as use those to normalise the data.

For this analysis we use `RUVSeq`

RUVg normalization is performed in several fundamental steps: 

Step 1: run a normal DGE 
  1.1: design of a model matrix (here we use the same design as the TMM normalization).
  
  1.2: A fit is then constructed using normalization factor (here we use TMM)
  
  1.3: perform GLMQLF test.
     note: here we have no contrasts and are instead performing an ANOVA-like test - this is what allows us to identify genes which do not vary between samples. 
     
Step 2: identify the DEG

Step 3: create a set of 5000 "empirical" genes with the highest PValue (ie. not changing). 
  note: this is somewhat arbitrary

Step 4: run RUVg to estimate the variation (where estimated factor of unwanted variation (k) = 1)


```{r ruvg-nomralisation}
library(RUVSeq)

# step 1: run a normal DGE 
  # 1.1: same model design as before.
model <- model.matrix(~ disease_state, data = colData(SE_outlier_removed))
# model

  # 1.2: same fit as before using TMM for normlaisation
fit <- SE_outlier_removed %>%
  calcNormFactors(method = "TMM") %>%
  estimateDisp(design = model) %>%
  glmQLFit(model)

# 1.3: perform GLMQLF test.
      # note: here we have no contrasts and are instead performing an ANOVA-like test  

QLFTest <- glmQLFTest(fit, coef = 2:3)
# QLFTest


# step 2: return the differential expressed genes.

QLFtop <- QLFTest %>%
  topTags(n = Inf) %>%
  pluck("table") %>%
  dplyr::select(
    starts_with("gene"),
    logCPM,
    'F',
    PValue,
    FDR)


#step 3: create a set of 5000 of genes with the highest PValue (ie. not changing). 
  # note: this is somewhat arbitrary

emperical <- QLFtop %>%
  dplyr::arrange(desc(PValue)) %>%
  .[1:5000,] %>%
  .$gene_id


# step 4: run RUVg (where estimated factor of unwanted variation (k) = 1)

ruvg <- RUVg(assay(SE_outlier_removed, "counts"), emperical, k = 1) 
ruvg$W

# due to compatibility issues  - convert to a DGElist object 
ruv_dge <- DGEList(
  counts = ruvg$normalizedCounts,
  samples = cbind(colData(SE_outlier_removed), ruvg$W),
  genes = rowRanges(SE_outlier_removed))

# use extraChIPs to plot data 
plotAssayRle(SE_outlier_removed, "RUVGnorm", fill = "disease_state") +
  scale_fill_brewer(palette = "Accent") + 
  theme_bw()


```

Here we can see that the mean relative log expression across all samples sits around 0. This indicates the normalization is effective. 

We then proceed to generate the fit and model from the subsequent DGEA.
As with the TMM normalisation alone, here we used the Quasi-liklihood method to generate the model fit.
```{r RUVG_fit_and_model}

# create a new design model including the estimated variation.

RUVg_model <- model.matrix(~ disease_state + W_1, data = ruv_dge$samples)


# create a new fit
  # note: We still use TMM normalisation here 

RUVg_fit <- ruv_dge %>%
  calcNormFactors(method = "TMM") %>%
  estimateDisp(RUVg_model) %>%
  glmQLFit(RUVg_model)

# create a new contrast matrix with the levels from model. 

RUVg_contrasts <- makeContrasts(
  T2D_v_DN = disease_statetype2_diabetes - disease_stateDiabetic_nephropathy,  
  Hv_v_T2D = disease_statetype2_diabetes,
  Hv_v_DN = disease_stateDiabetic_nephropathy,
  levels = RUVg_model)

```

## Health Control Vs. Type 2 Diabetes
As before we perform differential expression using the `glmTreat` function, with the Log2FC threshold set at 1.2.
here we show a brief summary of DEG's
```{r HV-vs-T2D}

RUV_Hv_V_T2D <- glmTreat(RUVg_fit, contrast = RUVg_contrasts[, "Hv_v_T2D"],  lfc = log2(1.2)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
    mutate(Significant = PValue < 0.05) %>%
    mutate(direction = case_when(
    Significant == TRUE & logFC > 0 ~ "Up",
    Significant == TRUE & logFC < 0 ~ "Down",
    Significant == FALSE ~ "NotSignificant")
    )

RUV_Hv_V_T2D %>%
  summarise(
    n = dplyr::n(),
    .by = direction
  ) %>% 
  dplyr::rename_with(str_to_title) %>% 
  pander(
    justify = "lr",
    caption = "Summary of significant DE genes by direction"
  )


```

### MA plot (Health Control Vs. Type 2 Diabetes)
```{r MA_RUV_Hv_V_T2D }
MA_RUV_Hv_V_T2D  <- RUV_Hv_V_T2D %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  theme_bw()+
  geom_point(aes(colour = direction,
                 text = paste(gene_name)
                 )) +
  scale_colour_manual(values = c("#BEAED4", "black", "#7FC97F")) +
  geom_smooth(method = "glm", colour = "#386CB0") +
  ggtitle("Health control vs Type 2 diabetes MA plot (RUVg normalisation)")
MA_RUV_Hv_V_T2D 

```

### Volcano plot (Health Control Vs. Type 2 Diabetes)
```{r RUV_Hv_V_T2D-Vol-plot}
vol_RUV_Hv_V_T2D  <- RUV_Hv_V_T2D %>%
  ggplot(aes(
    x = logFC,
    y = -log10(PValue)
  )) + 
  theme_bw()+
  geom_point(aes(colour = direction, 
                 alpha = logCPM, 
                 #size = logCPM, 
                 text = paste(gene_name)
                 )) +
  scale_colour_manual(values = c("#BEAED4", "black", "#7FC97F")) +
  scale_alpha(range = c(0.3, 1)) +
  #scale_size(range = c(1, 4)) + 
  ggtitle("Health Control Vs. Type 2 Diabetes volcano plot (RUVg normaisation)")
# ggplotly(vol_Hv_v_T2D_treat, tooltip = c("text", "logCPM"))

vol_RUV_Hv_V_T2D


```

### Table of DEGS (Healthy Control vs Type 2 Diabetes)
```{r RUV_Hv_V_T2D-DEG-summary}

cp <- htmltools::tags$caption(
  "Healthy Control vs Type 2 Diabetes"
)

tbl_ID <- "tbl-RUV-Hv-V-T2D"

tbl <- RUV_Hv_V_T2D %>%
  dplyr::filter(Significant == TRUE) %>%
  dplyr::select(-c("start", "end", "width", "detected")) %>%
  reactable(
    filterable = TRUE,
    columns = list(
      seqnames = colDef("Chr", maxWidth = 70),
      logFC = colDef(format = colFormat(digits = 2)),
      unshrunk.logFC = colDef(show = FALSE),
      Significant = colDef(show = FALSE),
      PValue = colDef(
        cell = \(value) format_p(value)
      ),
      FDR = colDef(
        cell = \(value) format_p(value)
      ),
      logCPM = colDef(
        cell = \(value) format_p(value))
    ),
    theme = reactableTheme(style = list(fontSize = 12)),
    elementId = tbl_ID
  )

browsable(
  tagList(
    tags$button(
      tagList(fontawesome::fa("download"), "Download as CSV"),
      onclick = glue::glue("Reactable.downloadDataCSV('{tbl_ID}', '{tbl_ID}.csv')")
    ),
    tags$br(), cp,
    tbl
  )
)


```

## Healthy control vs. Diabetic Nephropathy.

### Summary of DEG's
```{r HV-vs-DN}

RUV_Hv_V_DN <- glmTreat(RUVg_fit, contrast = RUVg_contrasts[, "Hv_v_DN"],  lfc = log2(1.5)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05) %>%
  mutate(direction = case_when(
    Significant == TRUE & logFC > 0 ~ "Up",
    Significant == TRUE & logFC < 0 ~ "Down",
    Significant == FALSE ~ "NotSignificant")
    )

RUV_Hv_V_DN %>%
  summarise(
    n = dplyr::n(),
    .by = direction
  ) %>% 
  dplyr::rename_with(str_to_title) %>% 
  pander(
    justify = "lr",
    caption = "Summary of significant DE genes by direction"
  )


```

### MA plot (Health Control Vs. Diabetic Nephropathy)
```{r MA_RUV_Hv_V_DN }
MA_RUV_Hv_V_DN <- RUV_Hv_V_DN %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
  )) + 
  theme_bw()+
  geom_point(aes(colour = direction,
                 text = paste(gene_name)
                 )) +
  scale_colour_manual(values = c("#BEAED4", "black", "#7FC97F")) +
  geom_smooth(method = "glm", colour = "#386CB0") +
  ggtitle("Health control vs Diabetic Nephropathy MA plot (RUVg normalisation)")
MA_RUV_Hv_V_DN 

```

### Volcano plot (Health Control Vs. Diabetic Nephropathy)

```{r RUV_Hv_V_DN-Vol-plot}
vol_RUV_Hv_V_DN  <- RUV_Hv_V_DN %>%
  ggplot(aes(
    x = logFC,
    y = -log10(PValue)
  )) + 
  theme_bw()+
  geom_point(aes(colour = direction, 
                 alpha = logCPM, 
                 #size = logCPM, 
                 text = paste(gene_name)
                 )) +
  scale_colour_manual(values = c("#BEAED4", "black", "#7FC97F")) +
  scale_alpha(range = c(0.3, 1)) +
  #scale_size(range = c(1, 4)) + 
  ggtitle("Health Control Vs. Diabetic Nephropathy volcano plot (RUVg normaisation)")
# ggplotly(vol_Hv_v_T2D_treat, tooltip = c("text", "logCPM"))

vol_RUV_Hv_V_DN


```

### Table of DEGS (Healthy Control vs Diabetic Nephropathy)
```{r RUV_Hv_V_DN-DEG-summary}

cp <- htmltools::tags$caption(
  "Healthy Control vs Diabetic Nephropathy"
)

tbl_ID <- "tbl-RUV-Hv-V-DN"

tbl <- RUV_Hv_V_DN %>%
  dplyr::filter(Significant == TRUE) %>%
  dplyr::select(-c("start", "end", "width", "detected")) %>%
  reactable(
    filterable = TRUE,
    columns = list(
      seqnames = colDef("Chr", maxWidth = 70),
      logFC = colDef(format = colFormat(digits = 2)),
      unshrunk.logFC = colDef(show = FALSE),
      Significant = colDef(show = FALSE),
      PValue = colDef(
        cell = \(value) format_p(value)
      ),
      FDR = colDef(
        cell = \(value) format_p(value)
      ),
      logCPM = colDef(
        cell = \(value) format_p(value))
    ),
    theme = reactableTheme(style = list(fontSize = 12)),
    elementId = tbl_ID
  )

browsable(
  tagList(
    tags$button(
      tagList(fontawesome::fa("download"), "Download as CSV"),
      onclick = glue::glue("Reactable.downloadDataCSV('{tbl_ID}', '{tbl_ID}.csv')")
    ),
    tags$br(), cp,
    tbl
  )
)


```

## Type 2 diabetes vs. Diabetic Nephropathy.

### Summary of DEG's
```{r T2D-vs-DN}

RUV_T2D_V_DN <- glmTreat(RUVg_fit, contrast = RUVg_contrasts[, "T2D_v_DN"],  lfc = log2(1.5)) %>%
topTags(n = Inf) %>%
  pluck("table") %>%
  as_tibble() %>%
  mutate(Significant = PValue < 0.05) %>%
  mutate(direction = case_when(
    Significant == TRUE & logFC > 0 ~ "Up",
    Significant == TRUE & logFC < 0 ~ "Down",
    Significant == FALSE ~ "NotSignificant")
    )

RUV_T2D_V_DN %>%
  summarise(
    n = dplyr::n(),
    .by = direction
  ) %>% 
  dplyr::rename_with(str_to_title) %>% 
  pander(
    justify = "lr",
    caption = "Summary of significant DE genes by direction"
  )


```

### MA plot (Tye 2 Diabetes Vs. Diabetic Nephropathy)
```{r MA_RUV_T2D_V_DN }
MA_RUV_T2D_V_DN <- RUV_T2D_V_DN %>%
  ggplot(aes(
    x = logCPM,
    y = logFC
    )
  ) + 
  theme_bw()+
  geom_point(aes(colour = direction,
                 text = paste(gene_name)
                 )) +
  scale_colour_manual(values = c("#BEAED4", "black", "#7FC97F")) +
  geom_smooth(method = "glm", colour = "#386CB0") +
  ggtitle("Type 2 diabetes vs Diabetic Nephropathy MA plot (RUVg normalisation)")
MA_RUV_T2D_V_DN 

```

### Volcano plot (Type 2 Diabetes Vs. Diabetic Nephropathy)

```{r RUV_T2D_V_DN-Vol-plot}
vol_RUV_T2D_V_DN  <- RUV_T2D_V_DN %>%
  ggplot(aes(
    x = logFC,
    y = -log10(PValue)
  )) + 
  theme_bw()+
  geom_point(aes(colour = direction, 
                 alpha = logCPM, 
                 #size = logCPM, 
                 text = paste(gene_name)
                 )) +
  scale_colour_manual(values = c("#BEAED4", "black", "#7FC97F")) +
  scale_alpha(range = c(0.3, 1)) +
  #scale_size(range = c(1, 4)) + 
  ggtitle("Type 2 Diabetes Vs. Diabetic Nephropathy volcano plot (RUVg normaisation)")
# ggplotly(vol_Hv_v_T2D_treat, tooltip = c("text", "logCPM"))

vol_RUV_T2D_V_DN


```

### Table of DEGS (Type 2 Diabetes vs Diabetic Nephropathy)
```{r RUV_T2D_V_DN-DEG-summary}

cp <- htmltools::tags$caption(
  "Type 2 Diabetes  vs Diabetic Nephropathy"
)

tbl_ID <- "tbl-RUV-T2D-V-DN"

tbl <- RUV_T2D_V_DN %>%
  dplyr::filter(Significant == TRUE) %>%
  dplyr::select(-c("start", "end", "width", "detected")) %>%
  reactable(
    filterable = TRUE,
    columns = list(
      seqnames = colDef("Chr", maxWidth = 70),
      logFC = colDef(format = colFormat(digits = 2)),
      unshrunk.logFC = colDef(show = FALSE),
      Significant = colDef(show = FALSE),
      PValue = colDef(
        cell = \(value) format_p(value)
      ),
      FDR = colDef(
        cell = \(value) format_p(value)
      ),
      logCPM = colDef(
        cell = \(value) format_p(value))
    ),
    theme = reactableTheme(style = list(fontSize = 12)),
    elementId = tbl_ID
  )

browsable(
  tagList(
    tags$button(
      tagList(fontawesome::fa("download"), "Download as CSV"),
      onclick = glue::glue("Reactable.downloadDataCSV('{tbl_ID}', '{tbl_ID}.csv')")
    ),
    tags$br(), cp,
    tbl
  )
)


```